Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-16 paper/main.tex
through 3.1

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  Together, $X$ and $\bar{Y}$ comprise the 
training data for a particular cosmology. $\bar{Y}$ is typically pre-processed
based on domain expertise, such as that defined in Section \ref{sec:data}.
We will reserve details of domain-specific pre-processing for the CAMB and Mira-Titan suites
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

@@

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we specify a dense covariance matrix $\Sigma_\varepsilon$ that simulataneously
allows us to model the correlated errors, as well as build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

We next place a Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\boldsymbol{\mu}_S, \Sigma_S(\cdot)\right)$, again with a 
Mat\'ern kernel. Here, we set $\boldsymbol{\mu}_S=\mathbf{0}$, although our model
is applicable to any selection of $\boldsymbol{\mu}_S$. Standard practice would 
use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$, yet we have found that 
prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$ via $W \sim \mathrm{GP}\left(\boldsymbol{\mu}_W, \Sigma_W(X)\right)$.
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
Again, we set $\boldsymbol{\mu}_W = \mathbf{0}$ without loss of generatlity (we have
found $\boldsymbol{\mu}_W = X$ works similarly).
Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right).
\end{aligned}
\end{equation}
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernel \textcolor{red}{@Annie, might need to add/remove $\tau^2$ 
in one or both of these kernels},
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2. 
% \;\;\textrm{and} \\
% &\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
%   \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
% \quad\textrm{for}\quad 
% d=||x_i-x_j||^2,
\end{aligned}
\]
In the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the correlation 
scale and lengthscale respectively.  In the $W$ layer, $\Sigma_W(X)^{i,j}$ is defined 
analogously for squared distances $d=||x_i-x_j||^2$ and we allow for a unique 
lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.

@@

We embrace fully-Bayesian inference of latent $W$ and kernel hyperparameters $\{\tau^2, \theta_S, \theta_W\}$.
Specifically, we use elliptical slice sampling \citep[ESS;][]{murray2010elliptical} to 
generate posterior samples of $W$, using Eq.~(\ref{eq:prior}) for proposal generation and 
Eq.~(\ref{eq:likelihood}) for likelihood-based acceptance.  
Under a reference prior $\pi(\tau^2)\propto\frac{1}{\tau^2}$,
we integrate $\tau^2$ from the likelihood of Eq.~(\ref{eq:likelihood}).  The resulting marginal 
posterior is equivalent to that with the plug-in estimator 
$\hat{\tau}^2 = \dots$ \citep[][Chapter 5]{gramacy2020surrogates}.
\textcolor{red}{(With $\Sigma_\epsilon$ on there, we can't actually do this.. unless we divide
$\Sigma_\epsilon$ by $\tau^2$...)}.
For $\theta_S$ and $\theta_W$, after pre-scaling, we adopt the default prior specifications 
provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
We then integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework 
with the ESS sampling of $W$ \citep{sauer2023active}.

>>>>

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  
%Together, $X$ and $\bar{Y}$ comprise the 
%training data for a particular cosmology. 
$\bar{Y}$ is typically pre-processed
based on domain expertise.  Pre-processing for Mira-Titan was described in Section 
\ref{sec:data}; we will detail domain-specific pre-processing 
for the CAMB suite in Section \ref{subsec:camb}.

@@

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence
(as visualized in Figure \ref{fig:plot_data}).  We thus specify a dense 
covariance matrix $\Sigma_\varepsilon$ that simulataneously
allows us to model the correlated errors and build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

We next place a Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\boldsymbol{\mu}_S, \Sigma_S(\cdot)\right)$, again with a 
Mat\'ern kernel. Here, we set $\boldsymbol{\mu}_S=\mathbf{0}$, although our model
is applicable to any selection of $\boldsymbol{\mu}_S$. Standard practice would 
use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$, yet we have found that 
prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate a latent monotonic Gaussian process \citep[monoGP;][]{barnett2025monotonic},
which accommodates nonstationarity by warping $X$ via 
$W \sim \mathrm{monoGP}\left(\boldsymbol{\mu}_W, \Sigma_W(X)\right)$.
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
Again, we set $\boldsymbol{\mu}_W = \mathbf{0}$ without loss of generatlity (we have
found $\boldsymbol{\mu}_W = X$ works similarly).
Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{monoGP}\left(\mathbf{0}, \Sigma_W(X)\right).
\end{aligned}
\end{equation}
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

For both $\Sigma_S$ and $\Sigma_W$, we use Mat\`ern kernels with unit scale and
unique lengthscales,
\[
\begin{array}{l}
\Sigma_S(W)^{i,j} = k\left(||w_i - w_j||^2, \theta_S\right) \\[7pt]
\Sigma_W(X)^{i,j} = k\left(||x_i - x_j||^2, \theta_W\right)
\end{array}
\quad\textrm{where}\quad
k(d, \theta) = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta}} + 
  \frac{5d^2}{3\theta}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta}}\right).
\]
In our one-dimensional setting, we embrace unit scale on $S$ to preserve identifiability
with $\theta_S$ \textcolor{red}{(@Steve - can you cite that paper here?)}.
Unit scale on latent $W$ was recommended by \citet{sauer2023active} for similar reasons.

@@

We embrace fully-Bayesian inference of latent $W$ and kernel hyperparameters $\{\theta_S, \theta_W\}$.
Specifically, we use elliptical slice sampling \citep[ESS;][]{murray2010elliptical} to 
generate posterior samples of $W$, using Eq.~(\ref{eq:prior}) for proposal generation and 
Eq.~(\ref{eq:likelihood}) for likelihood-based acceptance.  
Under a reference prior $\pi(\tau^2)\propto\frac{1}{\tau^2}$,
we integrate $\tau^2$ from the likelihood of Eq.~(\ref{eq:likelihood}).  The resulting marginal 
posterior is equivalent to that with the plug-in estimator 
$\hat{\tau}^2 = \dots$ \citep[][Chapter 5]{gramacy2020surrogates}.
\textcolor{red}{(With $\Sigma_\epsilon$ on there, we can't actually do this.. unless we divide
$\Sigma_\epsilon$ by $\tau^2$...)}.
For $\theta_S$ and $\theta_W$, after pre-scaling, we adopt the default prior specifications 
provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
We then integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework 
with the ESS sampling of $W$ \citep{sauer2023active}.
